Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises

### 2021-04-27 penguins.tex
Aph deux (#21)
* repurposed R/penguins_code.Rmd with penguins_plots.R to reduce duplication

* rm fig dir

* start keeping these files? not sure if we want to separate

* commit those files

* all png files now go here when you knit

* this is a purled script

* add tex file to git repo

* rework manuscript for better reproducibility

* don't show wizard behind curtain

* booktabs

* small edits

* updates

* no more gt?

* I don't even know

% !TeX root = RJwrapper.tex
\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package - An
Alternative to Anderson's \emph{Irises}}
\author{by Allison M. Horst, Alison P. Hill, Kristen B. Gorman}

@@

\abstract{%
Edgar Anderson's famous 1935 \emph{Iris} dataset, which exists in the
base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in
statistics and data science education. Due to a lack of metadata,
non-intuitive variables (e.g.~Sepal.Width), its publication in (and
inextricable connection with) Fisher's 1936 paper in \emph{Annals of
Eugenics}, and perfect structure (e.g.~no missing values)
unrepresentative of most real-world environmental data, a modern
alternative to \textbf{iris} is needed. Penguin morphological
measurements for three species of \emph{Pygoscelis} penguins that breed
on islands throughout the Palmer Archipelago, Antarctica, provide an
approachable, charismatic and near drop-in replacement for the
\textbf{iris} data. Here, we introduce the \textbf{palmerpenguins} R
package and describe the included \textbf{penguins} data. We directly
compare the two datasets for selected analyses to demonstrate that R
users, in particular teachers and learners currently using the
\textbf{iris} data, can switch to the Palmer Archipelago penguins for
many use cases including data wrangling and visualization, correlation,
regression, hypothesis testing, multivariate analysis (e.g.~PCA),
cluster analysis and classification (e.g.~by k-means).
}

@@

In 1935, American botanist Edgar Anderson measured petal and sepal
structural dimensions (length and width) for 50 flowers from three Iris
species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
size (5 variables and 150 total observations) and characteristics of the
data make it amenable for introducing a wide range of statistical
methods including data wrangling, visualization, regression,
multivariate exploration, and machine learning. Anderson's \emph{Iris}
dataset is built into a number of software packages including the
auto-installed datasets package in R \citep{r_core_team_r_2019},
Python's scikit-learn machine learning library
\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
Institute, Cary NC), which have facilitated widespread use of the data.
As a result, eighty-five years after the data were initially published,
Anderson's \emph{Iris} measurements are ubiquitous in statistics,
computational methods, and data science courses and materials.

There are several reasons that practitioners and educators may want to
retire Anderson's \emph{Iris} dataset in modern data science materials,
including:

\begin{itemize}
\tightlist
\item
  Anderson's \emph{Iris} dataset entirely lacks metadata, limiting
  meaningful interpretation and discussion of research methods,
  analyses, and outcomes
\item
  Two of its five variables, \texttt{Sepal.Width} and
  \texttt{Sepal.Length}, are not intuitive to most users
\item
  In 1936, the \emph{Iris} data were published by eugenicist R.A. Fisher
  in \emph{Annals of Eugenics}, and are now inextricably linked to
  Fisher's eugenics work; in fact, the dataset is commonly referred to
  as \emph{Fisher's Iris data}
\item
  The \emph{Iris} dataset contains only complete cases (i.e.~no missing
  values) and equal sample sizes (\emph{n} = 50 for each \emph{Iris}
  species) - more perfect than real-world datasets learners are likely
  to encounter ``in the wild''
\item
  The single categorical variable (\texttt{Species}) limits options for
  analyses
\end{itemize}

Given these motivations to retire \emph{Iris}, we sought to identify,
then make readily available, a suitable alternative dataset. Considering
\emph{Iris} usage in data science and statistics materials, we
established the following criteria for a suitable alternative:

\begin{itemize}
\tightlist
\item
  Available by appropriate license (ideally, CC0 ``no rights reserved'')
\item
  Features intuitive subjects and variables that are interesting and
  understandable to learners across fields
\item
  Complete metadata and documentation
\item
  Manageable (but not trivial) in size
\item
  Minimal data cleaning and pre-processing required for most analyses
\item
  Real-world (not manufactured), modern data
\item
  Similar opportunities for teaching and learning R, data science, and
  statistical skills
\item
  Can easily replace \emph{Iris} for most use cases
\end{itemize}

Here, we describe an alternative to Anderson's \emph{Iris} data that
largely satisfies these criteria: a refreshing, approachable and
charismatic dataset containing real-world morphological data for three
\emph{Pygoscelis} penguin species that breed throughout the Western
Antarctic Peninsula region, made available through the Long-Term
Ecological Research Network (US LTER). By comparing data structure,
size, and a range of analyses side-by-side for the two datasets, we
demonstrate that the Palmer Archipelago penguin measurements are an
ideal substitute for Anderson's \emph{Iris} data for many use cases in
statistics and data science education.

@@

Body size measurements, clutch (i.e., egg laying) observations (e.g.,
date of first egg laid, and clutch completion), and carbon
(\textsuperscript{13}C/\textsuperscript{12}C,
\(\delta\)\textsuperscript{13}C) and nitrogen
(\textsuperscript{15}N/\textsuperscript{14}N,
\(\delta\)\textsuperscript{15}N) stable isotope values of red blood
cells for male and female adult Adélie (\emph{P. adeliae}), chinstrap
(\emph{P. antarcticus}), and gentoo (\emph{P. papua}) penguins on three
islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were
collected from 2007 - 2009 by Dr.~Kristen Gorman in collaboration with
the \href{https://pal.lternet.edu/}{Palmer Station LTER}, part of the
\href{https://lternet.edu/}{US LTER Network}. For complete data
collection methods and published analyses see Gorman et al.~(2014).

The data in the \textbf{palmerpenguins} R package are available for use
by CC0 license (``No Rights Reserved'') in accordance with the
\href{https://pal.lternet.edu/data/policies}{Palmer Station LTER Data
Policy} and the \href{https://lternet.edu/data-access-policy/}{LTER Data
Access Policy}, and were imported from the
\href{https://environmentaldatainitiative.org/}{Environmental Data
Initiative (EDI) Data Portal} at the links below:

@@

\hypertarget{r-package-palmerpenguins}{%
\subsection{\texorpdfstring{R package:
\textbf{palmerpenguins}}{R package: palmerpenguins}}\label{r-package-palmerpenguins}}

R users can install the \textbf{palmerpenguins} package from CRAN by:

\begin{verbatim}
  `install.packages("palmerpenguins")`
\end{verbatim}

Alternatively, the development version of the package can be installed
from GitHub:

\begin{verbatim}
  `remotes::install_github("allisonhorst/palmerpenguins")`
\end{verbatim}

Information, examples, and links to community-contributed materials are
available on the \textbf{palmerpenguins} package website:
\url{https://allisonhorst.github.io/palmerpenguins/}.

The \textbf{palmerpenguins} R package contains two data objects:
\textbf{penguins\_raw} and \textbf{penguins}. The \textbf{penguins\_raw}
data consists of all raw data for 17 variables (Appendix Table 1),
recorded completely or in part for 344 individual penguins, accessed
directly from EDI. As a direct alternative to Anderson's \emph{Iris}
data we recommend using the curated data in \textbf{penguins}, which is
a subset of \textbf{penguins\_raw} retaining all 344 observations,
minimally updated (Appendix B) and reduced to the following eight
variables:

\begin{itemize}
\tightlist
\item
  \emph{species:} a factor denoting the penguin species (Adélie,
  chinstrap, or gentoo)
\item
  \emph{island:} a factor denoting the island (in Palmer Archipelago,
  Antarctica) where observed (Biscoe, Dream or Torgersen)
\item
  \emph{bill\_length\_mm:} a number denoting length of the dorsal ridge
  of a penguin bill (millimeters)
\item
  \emph{bill\_depth\_mm:} a number denoting the depth of a penguin bill
  (millimeters)
\item
  \emph{flipper\_length\_mm:} an integer denoting the length of a
  penguin flipper (millimeters)
\item
  \emph{body\_mass\_g:} an integer denoting the weight of a penguin's
  body (grams)
\item
  \emph{sex:} a factor denoting the sex of a penguin sex (male, female)
  based on molecular data
\item
  \emph{year:} an integer denoting the year of study (2007, 2008 or
  2009)
\end{itemize}

The same data exist as comma-separated value (CSV) files in the package
(``penguins\_raw.csv'' and ``penguins.csv''), and can be read in using
the built-in \texttt{path\_to\_file()} function in
\textbf{palmerpenguins}. For example,

\begin{verbatim}
    library(tidyverse)
    library(palmerpenguins)
    df <- read_csv(path_to_file(“penguins.csv”))
\end{verbatim}

will read in ``penguins.csv'' as if from an external file, thus
automatically parsing \emph{species}, \emph{island} and \emph{sex}
variables as characters. This option allows users opportunities to
practice or demonstrate reading in data from a CSV, then updating
variable class (e.g.~characters to factors).

\hypertarget{other-data-access-options}{%
\subsection{Other data access options}\label{other-data-access-options}}

Python: Python users can access the penguins data in the
\textbf{seaborn} data visualization library (Waskom et al.~2017).
Example code to load the data in Python:

Julia: Julia users can access the penguins data in the
\textbf{PalmerPenguins.jl} package. Example code to import the penguins
data through \textbf{PalmerPenguins.jl}:

\begin{verbatim}
    julia> using DataFrames
    julia> df = DataFrame(table)
\end{verbatim}

\hypertarget{software-and-code}{%
\subsection{Software and code}\label{software-and-code}}

All analyses were performed in the R language environment using version
3.6.2 (R Core Team 2019). Complete code for this paper is shared in the
Supplemental Material. We acknowledge the following R packages used in
analyses, with gratitude to developers and contributors:

\begin{itemize}
\tightlist
\item
  \textbf{tidyverse} (Wickham et al.~2019), for data import and cleaning
\item
  \textbf{ggplot2} (Wickham 2016:2), for data visualizations
\item
  \textbf{here} (Müller 2017), for file path control
\item
  \textbf{kableExtra} (Zhu 2019) for finalized tables
\item
  \textbf{gt} (Iannone, Cheng, and Schloerke 2020) for finalized tables
\item
  \textbf{GGally} (Schloerke et al.~2020), for pairs plots
\item
  \textbf{patchwork} (Pedersen 2019), for compound figures
\item
  \textbf{shadowtext} (Yu 2019), to add a background color to text
  labels
\item
  \textbf{recipes} (Kuhn and Wickham 2020), for data pre-processing
\item
  \textbf{base} and \textbf{stats} (R Core Team 2019) for various
  analyses throughout
\item
  \textbf{pkgdown} (ref) was used to build the package website
\end{itemize}

\hypertarget{selected-comparisons-between-iris-and-penguins}{%
\subsection{\texorpdfstring{Selected comparisons between \textbf{iris}
and
\textbf{penguins}}{Selected comparisons between iris and penguins}}\label{selected-comparisons-between-iris-and-penguins}}

The \textbf{penguins} data in \textbf{palmerpenguins} is generally
useful and approachable for data science and statistics education, and
is uniquely well-suited to replace the \textbf{iris} dataset.
Comparisons presented are selected examples for common \textbf{iris}
uses, and are not exhaustive.

@@

Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]}
with each column denoting a single variable and each row containing
measurements for a single \emph{Iris} flower or penguin. The two
datasets are comparable in size: dimensions (columns × rows) are 5 × 150
and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and
sample sizes within species are similar (Table 1). Notably, sample sizes
differ for the three penguin species, while sample sizes in
\textbf{iris} are equal (n = 50 for each Iris species). Multiple factor
variables in \textbf{penguins} (\emph{species}, \emph{island} and
\emph{sex}) along with \emph{year} create additional opportunities for
grouping, compared to the single factor (\emph{species}) in
\textbf{iris}.

Unlike \textbf{iris}, which contains only complete cases, the
\textbf{penguins} dataset contains a small number of missing values
(n\textsubscript{missing} = 19, out of 2,752 total values). Missing
values and unequal sample sizes are common in real-world data, and we
believe add learning value to \textbf{penguins}.

@@

\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{l|c|l|c|c|c}
\hline
\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cline{1-2} \cline{3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\hline
setosa & 50 & Adélie & 73 & 73 & 6\\
\hline
versicolor & 50 & chinstrap & 34 & 34 & 0\\
\hline
virginica & 50 & gentoo & 58 & 61 & 5\\
\hline
\end{tabular}
\end{table}

@@

Distributions, relationships between variables and clustering can be
visually explored between species for the four structural size
measurements in \textbf{penguins} (flipper length, body mass, bill
length and depth; Figure \ref{fig:penguins-pairs}) and \textbf{iris}
(sepal width and length, petal width and length; Figure
\ref{fig:iris-pairs}).

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:penguins-pairs}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:iris-pairs}
\end{figure}
\end{Schunk}

There exist numerous opportunities to explore linear relationships and
correlations in \textbf{penguins} and \textbf{iris} both within and
across species (Figures \ref{fig:penguins-pairs} \&
\ref{fig:iris-pairs}). Here, we highlight one comparison that is
uniquely similar: petal dimensions (petal length versus petal width)
from \textbf{iris}, and penguin size (flipper length versus body mass)
in \textbf{penguins} (\ref{fig:linear-example}). The overall trend
across all three species is approximately linear for both \textbf{iris}
and \textbf{penguins}. Teachers may encourage students to explore how
simple linear regression results and predictions differ when the species
variable is omitted, compared to multiple linear regression with species
included (\ref{fig:linear-example}).

@@

{\centering \includegraphics[width=6in]{fig/linear_example}

@@

\caption[Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B)]{Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B). Within-species linear model is visualized for each penguin or iris species.}\label{fig:linear-example}
\end{figure}
\end{Schunk}

Notably, distinctions between species are clearer for iris petals -
particularly, the much smaller petals for \emph{Iris setosa} - compared
to penguins, in which Adélie and chinstrap penguins are largely
overlapping in body size (body mass and flipper length), and are both
generally smaller than gentoos.

Simpson's Paradox is a data phenomenon in which a trend observed between
variables is reversed when data are pooled, omitting a meaningful
variable. While often taught and discussed in statistics courses,
finding a real-world and approachable example of Simpson's paradox can
be a challenge. Here, we show one (of several possible; see Figure
\ref{fig:penguins-pairs}) Simpson's Paradox examples in
\textbf{penguins}: exploring bill dimensions with and without
\emph{species} included (Figure \ref{fig:simpsons}). When penguin
\emph{species} is omitted (Figure \ref{fig:simpsons} A), bill length and
depth appear negatively correlated overall. The trend is reversed when
\emph{species} is included, revealing an obviously positive correlation
between bill length and bill depth within species (Figure
\ref{fig:simpsons} B).

@@

{\centering \includegraphics[width=6in]{fig/simpson_gg}

@@

\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox. Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}\label{fig:simpsons}
\end{figure}
\end{Schunk}

@@

Principal component analysis (PCA) is a dimensional reduction method
commonly used to explore patterns in multivariate data. PCA tutorials
frequently employ the \textbf{iris} dataset, which is useful for
teaching the method due to multivariate normality, and clear,
approachable outcomes for variable loadings and clustering.

>>>>

% !TeX root = RJwrapper.tex
\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package
- An Alternative to Anderson's Irises}
\author{by Allison M. Horst, Alison Presmanes Hill, and Kristen B. Gorman}

@@

\abstract{%
In 1935, Edgar Anderson collected size measurements for 150 flowers from
three species of \emph{Iris} on the Gaspé Peninsula in Quebec, Canada.
Since then, Anderson's \emph{Iris} observations have become a classic
dataset in statistics, machine learning, and data science teaching
materials. It is included in the base R \textbf{datasets} package as
\textbf{iris}, making it easy for users to access without knowing much
about it. However, the lack of data documentation, presence of
non-intuitive variables (e.g.~``sepal width''), and perfectly balanced
groups with zero missing values make \textbf{iris} an inadequate and
stale dataset for teaching and learning modern data science skills.
Users would benefit from working with a more representative, real-world
environmental dataset with a clear link to current scientific research.
Importantly, Anderson's \emph{Iris} dataset appeared in a 1936
publication by R. A. Fisher in the \emph{Annals of Eugenics} (which is
often the first-listed citation for the dataset), inextricably linking
\textbf{iris} to racist research and theory. Thus, a modern alternative
to the \textbf{iris} dataset is needed. In this paper, we introduce the
\textbf{palmerpenguins} R package, which includes morphological
measurements collected from 2007 - 2009 for three species of
\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
Archipelago, Antarctica. The \textbf{penguins} data provides an
approachable, charismatic, and near drop-in replacement for
\textbf{iris} with topical relevance global climate change and
environmental impacts on marine predators. Since the release on CRAN in
July 2020, the \textbf{palmerpenguins} package has been downloaded over
102,000 times, highlighting the demand and widespread adoption of this
viable \textbf{iris} alternative. We directly compare the \textbf{iris}
and \textbf{penguins} datasets for selected analyses to demonstrate that
R users, in particular teachers and learners currently using
\textbf{iris}, can switch to the Palmer Archipelago penguins for many
use cases including data wrangling, visualization, linear modeling,
multivariate analysis (e.g., PCA), cluster analysis and classification
(e.g., by k-means).
}

@@

In 1935, American botanist Edgar Anderson measured petal and sepal
structural dimensions (length and width) for 50 flowers from three
\emph{Iris} species: \emph{Iris setosa}, \emph{Iris versicolor}, and
\emph{Iris virginica} \citep{anderson_irises_1935}. The manageable but
non-trivial size (5 variables and 150 total observations) and
characteristics of the dataset, including linear relationships and
multivariate normality, have made the dataset amenable for introducing a
wide range of statistical methods including data wrangling,
visualization, linear modeling, multivariate analyses, and machine
learning. Anderson's \emph{Iris} data is built into a number of software
packages including the auto-installed datasets package in R
\citep{r_core_team_r_2019}, Python's scikit-learn machine learning
library \citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library
(SAS Institute, Cary NC), which have facilitated widespread use of the
\emph{Iris} data. As a result, eighty-six years after the data were
initially published, Anderson's \emph{Iris} measurements are ubiquitous
in statistics, computational methods, software documentation, and data
science courses and materials.

There are a number of reasons that modern data science practitioners and
educators may want to move on from Anderson's \emph{Iris} dataset.
First, Anderson's \emph{Iris} dataset lacks metadata almost entirely
\citep{anderson_irises_1935}, which does not reinforce best practices
and limits meaningful interpretation and discussion of research methods,
analyses, and outcomes. Of the five variables in the dataset, two
(\texttt{Sepal.Width} and \texttt{Sepal.Length}) are not intuitive for
most non-botanists. Even with explanation, the difference between
\emph{Iris} \emph{petal} and \emph{sepal} dimensions is not obvious.
Second, the \emph{Iris} dataset contains equal sample sizes for each of
the three species (\emph{n} = 50) with no missing values, which is
cleaner than most real-world datasets that learners are likely to
encounter ``in the wild.'' Third, the single factor (\texttt{Species})
in the \emph{Iris} data limits options for analyses. Finally, due to its
publication in the \emph{Annals of Eugenics} by eugenicist and
statistician R.A. Fisher \citep{fisher_use_1936}, the \emph{Iris}
dataset is burdened by a history in scientific racism, which we are
committed to addressing through the development of new data science
education products as described below.

Given the growing need for fresh data science-ready datasets, we sought
to identify an alternative dataset that could be made easily accessible
for a broad audience. After evaluating the positive and negative
features of the \emph{Iris} dataset in data science and statistics
materials, we established the following criteria for a suitable
alternative:

\begin{itemize}
\tightlist
\item
  Available by appropriate license (ideally, CC0 ``no rights reserved'')
\item
  Feature intuitive subjects and variables that are interesting and
  understandable to learners across disciplines
\item
  Complete metadata and documentation
\item
  Manageable (but not trivial) in size
\item
  Minimal data cleaning and pre-processing required for most analyses
\item
  Real-world (not manufactured) modern data
\item
  Provides similar opportunities for teaching and learning R, data
  science, and statistical skills
\item
  Can easily replace \emph{Iris} for most use cases
\end{itemize}

Here, we describe an alternative to Anderson's \emph{Iris} data that
largely satisfies these criteria: a refreshing, approachable, and
charismatic dataset containing real-world morphological data for three
\emph{Pygoscelis} penguin species that breed throughout the Western
Antarctic Peninsula region, made available through the United States
Long-Term Ecological Research (US LTER) Network. By comparing data
structure, size, and a range of analyses side-by-side for the two
datasets, we demonstrate that the Palmer Archipelago penguin data are an
ideal substitute for Anderson's \emph{Iris} dataset for many use cases
in statistics and data science education.

@@

Body size measurements, clutch (i.e., egg laying) observations (e.g.,
date of first egg laid, and clutch completion), and carbon
(\textsuperscript{13}C/\textsuperscript{12}C,
\(\delta\)\textsuperscript{13}C) and nitrogen
(\textsuperscript{15}N/\textsuperscript{14}N,
\(\delta\)\textsuperscript{15}N) stable isotope values of red blood
cells for adult male and female Adélie (\emph{P. adeliae}), chinstrap
(\emph{P. antarcticus}), and gentoo (\emph{P. papua}) penguins on three
islands (Biscoe, Dream and Torgersen) within the Palmer Archipelago were
collected from 2007 - 2009 by Dr.~Kristen Gorman in collaboration with
the \href{https://pal.lternet.edu/}{Palmer Station LTER}, part of the
\href{https://lternet.edu/}{US LTER Network}. For complete data
collection methods and published analyses, see
\citet{gorman_ecological_2014}. Throughout this paper, penguins species
referred to as ``Adélie'', ``Chinstrap'', and ``Gentoo.''

The data in the \CRANpkg{palmerpenguins} R package are available for use
by CC0 license (``No Rights Reserved'') in accordance with the
\href{https://pal.lternet.edu/data/policies}{Palmer Station LTER Data
Policy} and the \href{https://lternet.edu/data-access-policy/}{LTER Data
Access Policy}, and were imported from the
\href{https://environmentaldatainitiative.org/}{Environmental Data
Initiative (EDI) Data Portal} at the links below:

@@

\hypertarget{r-package-palmerpenguins}{%
\subsection{R package: palmerpenguins}\label{r-package-palmerpenguins}}

R users can install the \CRANpkg{palmerpenguins} package from CRAN:

\begin{verbatim}
install.packages("palmerpenguins")
\end{verbatim}

Information, examples, and links to community-contributed materials are
available on the \CRANpkg{palmerpenguins} package website:
\url{https://allisonhorst.github.io/palmerpenguins/}. See the Appendix
for how Python and Julia users can access the same data.

The \CRANpkg{palmerpenguins} R package contains two data objects:
\textbf{penguins\_raw} and \textbf{penguins}. The \textbf{penguins\_raw}
data consists of all raw data for 17 variables, recorded completely or
in part for 344 individual penguins, accessed directly from EDI
(\textbf{penguins\_raw} properties are summarized in Appendix B). We
generally recommend using the curated data in \textbf{penguins}, which
is a subset of \textbf{penguins\_raw} retaining all 344 observations,
minimally updated (Appendix A) and reduced to the following eight
variables:

\begin{itemize}
\tightlist
\item
  \emph{species:} a factor denoting the penguin species (Adélie,
  Chinstrap, or Gentoo)
\item
  \emph{island:} a factor denoting the Palmer Archipelago island in
  Antarctica where each penguin was observed (Biscoe Point, Dream
  Island, or Torgersen Island)
\item
  \emph{bill\_length\_mm:} a number denoting length of the dorsal ridge
  of a penguin bill (millimeters)
\item
  \emph{bill\_depth\_mm:} a number denoting the depth of a penguin bill
  (millimeters)
\item
  \emph{flipper\_length\_mm:} an integer denoting the length of a
  penguin flipper (millimeters)
\item
  \emph{body\_mass\_g:} an integer denoting the weight of a penguin's
  body (grams)
\item
  \emph{sex:} a factor denoting the sex of a penguin sex (male, female)
  based on molecular data
\item
  \emph{year:} an integer denoting the year of study (2007, 2008, or
  2009)
\end{itemize}

\newpage

The same data exist as comma-separated value (CSV) files in the package
(``penguins\_raw.csv'' and ``penguins.csv''), and can be read in using
the built-in \texttt{path\_to\_file()} function in
\CRANpkg{palmerpenguins}. For example,

\begin{verbatim}
    library(palmerpenguins)
    df <- read.csv(path_to_file("penguins.csv"))
\end{verbatim}

will read in ``penguins.csv'' as if from an external file, thus
automatically parsing \emph{species}, \emph{island}, and \emph{sex}
variables as characters instead of factors. This option allows users
opportunities to practice or demonstrate reading in data from a CSV,
then updating variable class (e.g., characters to factors).

\hypertarget{comparing-iris-and-penguins}{%
\subsection{\texorpdfstring{Comparing \textbf{iris} and
\textbf{penguins}}{Comparing iris and penguins}}\label{comparing-iris-and-penguins}}

The \textbf{penguins} data in \CRANpkg{palmerpenguins} is useful and
approachable for data science and statistics education, and is uniquely
well-suited to replace the \textbf{iris} dataset. Comparisons presented
are selected examples for common \textbf{iris} uses, and are not
exhaustive.

\caption{\label{tab:overview-tbl}Overview comparison of penguins and iris dataset features and characteristics.}
\centering
\begin{tabular}[t]{lcl}
\toprule
Feature & iris & penguins\\
\midrule
Year(s) collected & 1935 & 2007 - 2009\\
Dimensions (col x row) & 5 x 150 & 8 x 344\\
Documentation & minimal & complete metadata\\
Variable classes & double (4), factor (1) & double (2), int (3), factor (3)\\
Missing values? & no (n = 0; 0.0\%) & yes (n = 19; 0.7\%)\\
\bottomrule
\end{tabular}
\end{table}

@@

Both \textbf{iris} and \textbf{penguins} are in tidy format
\citep{wickham_tidy_2014} with each column denoting a single variable
and each row containing measurements for a single \emph{Iris} flower or
penguin, respectively. The two datasets are comparable in size:
dimensions (columns × rows) are 5 × 150 and 8 × 344 for \textbf{iris}
and \textbf{penguins}, respectively, and sample sizes within species are
similar (Table 1).

Notably, while sample sizes in \textbf{iris} across species are all the
same, sample sizes in \textbf{penguins} differ across the three species.
The inclusion of three factor variables in \textbf{penguins}
(\emph{species}, \emph{island}, and \emph{sex}), along with \emph{year},
create additional opportunities for grouping, faceting, and analysis
compared to the single factor (\emph{Species}) in \textbf{iris}.

Unlike \textbf{iris}, which contains only complete cases, the
\textbf{penguins} dataset contains a small number of missing values
(n\textsubscript{missing} = 19, out of 2752 total values). Missing
values and unequal sample sizes are common in real-world data, and adds
learning opportunities to the \textbf{penguins} dataset.

@@

\caption{\label{tab:counts-tbl}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{lclccc}
\toprule
\multicolumn{2}{c}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\midrule
setosa & 50 & Adélie & 73 & 73 & 6\\
versicolor & 50 & Chinstrap & 34 & 34 & 0\\
virginica & 50 & Gentoo & 58 & 61 & 5\\
\bottomrule
\end{tabular}
\end{table}

@@

Distributions, relationships between variables, and clustering can be
visually explored between species for the four structural size
measurements in \textbf{penguins} (flipper length, body mass, bill
length and depth; Figure \ref{fig:penguin-pairs}) and \textbf{iris}
(sepal width and length, petal width and length; Figure
\ref{fig:iris-pairs}).

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figs/penguin-pairs-1}

\caption[Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species]{Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.}\label{fig:penguin-pairs}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figs/iris-pairs-1}

\caption[Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.}\label{fig:iris-pairs}
\end{figure}
\end{Schunk}

Both \textbf{penguins} and \textbf{iris} offer numerous opportunities to
explore linear relationships and correlations, within and across species
(Figures \ref{fig:penguin-pairs} \& \ref{fig:iris-pairs}). A bivariate
scatterplot made with the \textbf{iris} dataset reveals a clear linear
relationship between petal length and petal width. Using
\textbf{penguins} (\ref{fig:linear-example}), we can create a uniquely
similar scatterplot with flipper length and body mass. The overall trend
across all three species is approximately linear for both \textbf{iris}
and \textbf{penguins}. Teachers may encourage students to explore how
simple linear regression results and predictions differ when the species
variable is omitted, compared to multiple linear regression with species
included (\ref{fig:linear-example}).

@@

{\centering \includegraphics[width=1\linewidth]{figs/linear-example-1}

@@

\caption[Representative linear relationships for (A)]{Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises. Within-species linear model is visualized for each penguin or iris species.}\label{fig:linear-example}
\end{figure}
\end{Schunk}

Notably, distinctions between species are clearer for iris petals -
particularly, the much smaller petals for \emph{Iris setosa} - compared
to penguins, in which Adélie and Chinstrap penguins are largely
overlapping in body size (body mass and flipper length), and are both
generally smaller than Gentoo penguins.

Simpson's Paradox is a data phenomenon in which a trend observed between
variables is reversed when data are pooled, omitting a meaningful
variable. While often taught and discussed in statistics courses,
finding a real-world and approachable example of Simpson's Paradox can
be a challenge. Here, we show one (of several possible - see Figure
\ref{fig:penguin-pairs}) Simpson's Paradox examples in
\textbf{penguins}: exploring bill dimensions with and without species
included (Figure \ref{fig:simpsons}). When penguin species is omitted
(Figure \ref{fig:simpsons} A), bill length and depth appear negatively
correlated overall. The trend is reversed when species is included,
revealing an obviously positive correlation between bill length and bill
depth within species (Figure \ref{fig:simpsons} B).

@@

{\centering \includegraphics[width=1\linewidth]{figs/simpsons-1}

@@

\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox. Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}\label{fig:simpsons}
\end{figure}
\end{Schunk}

@@

Principal component analysis (PCA) is a dimensional reduction method
commonly used to explore patterns in multivariate data. The
\textbf{iris} dataset frequently appears in PCA tutorials due to
multivariate normality and clear interpretation of variable loadings and
clustering.
